Distinct communities of snow algae bloom at different elevations

### 2019-10-20 doc/Casey1.Rmd
re-wrote intro thru line 85

Snow algae are cold-adapted microalgae that bloom in melting snow [@Leya2013]. Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017]. Loss of snow cover can lead to snow drought [@Staudinger2014;@Barnhart2016], and upon meltout exposes darker surfaces below which contributes to snow-albedo feedback [@Dery2007]. Thus, snow algae could play an important role in cryosphere regulation, yet the species composition within snow algae blooms remains poorly studied. Snow algae are the primary producers in a microbiome that includes grazers and heterotrophic fungi and bacteria (ref), and can be considered ecosystem engineers through their impacts on snow melt and by supplying carbon to this poorly understood microbiome. Knowing the identity and community composition of the algae themselves is of primary importance.

Until recently, snow algae species were defined and identified on the basis of morphology. Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]; conversely, different species can look nearly identical. In the lab cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes. Different species were thought to have diagnostic traits such as pigments or cell walls, but genetic studies show that morphological similarity may be misleading in this genus; for example, cell types formerly known as *Chloromonas cf. nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
<!-- based on this paragraph, I would not trust the morphology categories in my thing. Ill take em out anyway, but... chlainomonas is distinct, chloro distinct -->

<!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*. Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin (refs). *Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf. nivalis* [@Prochazkova2018c]. *Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter). Two species are described on the basis of morphology, although analysis of the genetic differences between these species is lacking. Although *Chlainomonas* are morphologically distinct from all known *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade. *Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019].

<!-- how communities vary from patch to patch. -->

@@

Whether different species of snow algae are restricted to a particular microhabitat within the melting snowpack remains unresolved. Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as water and fertilizer application increased cell counts of *Chlamydomonas cf. nivalis* in experimental manipulation plots in Alaska [@Ganey2017]. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. Two studies using microscopy found that red snow containing *Chlamydomonas cf. nivalis* to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in the subalpine zone below treeline [@Hoham1979a;@Nedbalova2008a]. @Ganey2018 hypothesized the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to snow with higher moisture content; however red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c].

Microscopic observation of snow algae blooms shows snow algae blooms can be either morphologically diverse [@Lutz2018] or highly similar [@Remias2013]. 18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a]. 18S is highly conserved and therefore suited to survey broadly across a variety of taxa (ref); however due to its high similarity between closely related species it is poorly suited to differentiating between closely related taxa. The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.

In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS. Our aims were to: 1) see what species were present, 2) compare community composition between samples, 3) and see how they are distributed across the landscape.

>>>>

Snow algae are cold-adapted microalgae that bloom in melting summer snow [@Leya2013]. Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017]. Depleted snowpack threatens water supplies and can lead to drought [@Staudinger2014;@Barnhart2016], and earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007]; thus, snow algae could play an important role in snow melt regulation. Furthermore, snow algae are the primary producers in a microbiome that includes grazers, fungi, and bacteria [@Hoham2001], but whether there are ecological distinctions between types of snow algae blooms has been poorly studied. The first step in understanding this microbiome is defining the algae species that are present in snow algae blooms, and their community structure.

Until recently, snow algae species were defined and identified on the basis of morphology. Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]. In the lab, cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes. Conversely, different genetic species can look nearly identical: for example, cell types formerly known as *Chloromonas cf. nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
<!-- based on this paragraph, I would not trust the morphology categories in my thing. Ill take em out anyway, but... chlainomonas is distinct, chloro distinct -->

<!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*. Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin [@Remias2005;Remias2016]. *Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf. nivalis* [@Prochazkova2018c]. *Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter). Although *Chlainomonas* look very different from *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade [@Novis2008]. *Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019]; in contrast, only two species are currently described in *Chlainomonas* [@Novis2008].
<!-- how communities vary from patch to patch. -->

@@

Snow algae grow in the liquid water surrounding rounded snow granules [@Leya2013]. In this micro-habitat, temperature is nearly constant, but snow moisture, pH, nutrients, and light can differ greatly within the snowpack depending on location of snow and time of year [@Loranger1988;@Kuhn2001;@Gorton2007;@Ganey2017]. Differences in snow algae community composition could offer insight into the environmental factors that determine snow algae distribution. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. Two studies using microscopy found that red snow containing *Chlamydomonas cf. nivalis* was prevalent in the above timberline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in conifer forests below timberline [@Hoham1979a;@Nedbalova2008a]. Differences in snow moisture could influence which species of snow algae grow; several species of snow algae are anecdotally reported to only inhabit very saturated, waterlogged snow [@Lutz2015;@Remias2013;@Novis2008;@Prochazkova2018c], while  *Chlamydomonas cf. nivalis* is hypothesized to be specially adapted to well-drained, drier snow [@Ganey2018].

<!-- Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as both water and NPK fertilizer increased cell counts of *Chlamydomonas cf. nivalis* in one study in Alaska . -->

<!-- im not doing network analysis, this is community structure -->
Microscopic observation shows that snow algae blooms can contain a diversity of cell morphologies in the same sample [@Lutz2018], but the genetic community structure of snow algae may not correlate with what is seen under the microscope. 18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a]. 18S is highly conserved and therefore suited to survey broadly across a variety of taxa; however, due to its high similarity between closely related species short segments of thie gene are poorly-differentiated among closely related taxa (ref). The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.

In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS. Our aims were to: 1) see what species were present, 2) compare community composition between samples, 3) and see how they are distributed across the landscape. We found ...
